---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

# Reading in datasets and loading packages:

```{r, warning=FALSE, message=FALSE}
# Loading packages
library("ggplot2")
library('vegan')
library('dplyr')

# Setting working directory
setwd("C:/Users/ebatz/Box Sync/Eviner lab shared/Evan/Research Projects/Julia BetaDiv")

# load data
data<-read.csv("Vernal_data_final.csv", stringsAsFactors=FALSE)
data$Year<-as.factor(data$Year)

# Remove bare ground
data <- data %>%
  select(-(BarGr))

# Group by zone
by.zone <- data %>% 
  group_by(Site.ID, Grazing, Year, Zone) %>% 
  summarise_if(is.numeric, mean)

# Group by pool
by.pool <- data %>% 
  group_by(Site.ID, Grazing, Year) %>% 
  summarise_if(is.numeric, mean)
```

__Visualization__

# NMDS Plot of Zones:

```{r}

# What level do we want to plot?
data_adonis = by.zone

# Creating a new variable of both zone and grazing
data_adonis$GrazingZone = paste(data_adonis$Grazing, data_adonis$Zone)

# Runs NMDS with 2 dimensions
nmds_run = metaMDS(data_adonis[,6:(ncol(data_adonis) - 1)],
               try = 50,
               trymax = 100,
               trace = FALSE,
               distance = "bray",
               wascores = TRUE
               )

# Sets the group variable to factor
data_adonis$group = as.factor(data_adonis$GrazingZone)

# Adds group column to dataframe of NMDS points
NMDS = bind_cols(data.frame(nmds_run$points), group = as.factor(c(data_adonis$group)))

# Calculates the midpoint of each group
NMDS.mean=aggregate(NMDS[,1:2], list(group=NMDS$group),mean)

# Calculating the shape of the ellipses
veganCovEllipse<-function (cov, center = c(0, 0), scale = 1, npoints = 100){
    theta <- (0:npoints) * 2 * pi/npoints
    Circle <- cbind(cos(theta), sin(theta))
    t(center + scale * t(Circle %*% chol(cov)))
}

# Creating blank dataframe to store ellipses
df_ell <- data.frame()
for(g in levels(NMDS$group)){
  df_ell <- rbind(df_ell, cbind(as.data.frame(with(NMDS[NMDS$group==g,],
                  veganCovEllipse(cov.wt(cbind(MDS1,MDS2),
                                          wt=rep(1/length(MDS1),length(MDS1)))$cov,
                                  center=c(mean(MDS1),mean(MDS2)))))
                  ,group=g))
}

# Adding other details for facetting
NMDS = right_join(NMDS, bind_cols(group = as.character(as.numeric(data_adonis$group)), data_adonis[,2:4]),
                      by = "group")

NMDS.mean = inner_join(NMDS.mean, bind_cols(group = as.character(as.numeric(data_adonis$group)), data_adonis[,2:4]),
                      by = "group")
NMDS.mean = NMDS.mean[!duplicated(NMDS.mean[,2:5]),]

df_ell = right_join(df_ell, bind_cols(group = as.character(as.numeric(data_adonis$group)), data_adonis[,2:4]),
                      by = "group")
```

Creating dataframe of species scores to use in plotting vectors:   

Variables here are:  

* MDS1 (Axis 1 score)
* MDS2 (Axis 2 score)
* Spec (species name)
* dist (total distance from the center)
* rank (ranking based on distance)

```{r}
for_scores <-  data_adonis[,6:(ncol(data_adonis) - 2)]
for_scores <- for_scores[,colSums(for_scores) / nrow(for_scores) > 2]

NMDS.spec = data.frame(nmds_run$species)
NMDS.spec$spec = rownames(NMDS.spec)
NMDS.spec$dist = apply(NMDS.spec[,1:2], 1, function(x) as.numeric(sqrt(sum(x^2))))
NMDS.spec$rank = rank(-NMDS.spec$dist)

# Dataframe of species scores
NMDS.spec
```

Edit the parameters here to make the figure prettier:

```{r}
# How many species to keep?
maxrank = 10

# Producing figure
ggplot(data = NMDS, aes(MDS1, MDS2)) + 
  geom_point(aes(shape = Zone), alpha = .5) +
  scale_shape(solid = FALSE) + 
  geom_path(data = df_ell, aes(x=MDS1, y=MDS2, group = Zone), size=1, linetype=1, alpha = .5) +
  geom_text(data = NMDS.mean, aes(label = NMDS.mean$Zone)) + 
  facet_wrap(~Grazing) +
  #geom_segment(data = NMDS.spec[NMDS.spec$rank <= maxrank,], 
  #             aes(x = 0, y = 0, xend = MDS1, yend = MDS2),
  #                arrow = arrow(length = unit(0.03, "npc"))) + 
  geom_text(data = NMDS.spec[NMDS.spec$rank <= maxrank,], 
               aes(x = MDS1, y = MDS2, label = spec)) +
  theme_bw()
```

Checking that this represents the default image well (red are species without labels):

```{r}
plot(nmds_run)
```

# NMDS plot for combined pools:

```{r, warning=FALSE, message=FALSE}
library(tidyverse); library(testthat)

data_adonis = by.pool

nmds_run = metaMDS(data_adonis[,5:ncol(data_adonis)],
               try = 50,
               trymax = 100,
               trace = FALSE,
               distance = "bray",
               binary = TRUE,
               wascores = TRUE
               )

data_adonis$group = as.factor(paste(data_adonis$Grazing, data_adonis$Year))
NMDS = bind_cols(data.frame(nmds_run$points), group = as.factor(c(data_adonis$group)))
NMDS.mean=aggregate(NMDS[,1:2], list(group=NMDS$group),mean)

# Calculating the shape of the ellipses
veganCovEllipse<-function (cov, center = c(0, 0), scale = 1, npoints = 100){
    theta <- (0:npoints) * 2 * pi/npoints
    Circle <- cbind(cos(theta), sin(theta))
    t(center + scale * t(Circle %*% chol(cov)))
}

df_ell <- data.frame()
for(g in levels(NMDS$group)){
  df_ell <- rbind(df_ell, cbind(as.data.frame(with(NMDS[NMDS$group==g,],
                  veganCovEllipse(cov.wt(cbind(MDS1,MDS2),
                                          wt=rep(1/length(MDS1),length(MDS1)))$cov,
                                  center=c(mean(MDS1),mean(MDS2)))))
                  ,group=g))
}

# Adding other details for facetting
NMDS = right_join(NMDS, bind_cols(group = as.character(as.numeric(data_adonis$group)), data_adonis[,2:3]),
                      by = "group")

NMDS.mean = right_join(NMDS.mean, bind_cols(group = as.character(as.numeric(data_adonis$group)), data_adonis[,2:3]),
                      by = "group")

NMDS.mean = NMDS.mean[!duplicated(NMDS.mean[,2:5]),]

NMDS.mean = NMDS.mean[order(NMDS.mean$Year),]

df_ell = right_join(df_ell, bind_cols(group = as.character(as.numeric(data_adonis$group)), data_adonis[,2:3]),
                      by = "group")
```

Creating dataframe of species scores to use in plotting vectors:

```{r}
for_scores <-  data_adonis[,5:ncol(data_adonis) - 1]
for_scores <- for_scores[,colSums(for_scores) / nrow(for_scores) > 5]
NMDS.spec = data.frame(nmds_run$species)
NMDS.spec$spec = rownames(NMDS.spec)
NMDS.spec$dist = apply(NMDS.spec[,1:2], 1, function(x) as.numeric(sqrt(sum(x^2))))
NMDS.spec$rank = rank(-NMDS.spec$dist)
NMDS.spec
```

Edit the parameters here to make the figure prettier:

```{r}
# How many species to plot?
maxrank = 10

# Producing figure
ggplot(data = NMDS, aes(MDS1, MDS2)) + 
  geom_point(aes(shape = Grazing), alpha = .5) +
  scale_shape(solid = FALSE) + 
  geom_path(data = df_ell, aes(x=MDS1, y=MDS2, group = Grazing), size=1, linetype=1) +
  geom_text(data = NMDS.mean, aes(label = NMDS.mean$Grazing)) +
  theme_bw() +
  facet_wrap(~Year) +
  #geom_segment(data = NMDS.spec[NMDS.spec$rank <= maxrank,], 
  #             aes(x = 0, y = 0, xend = MDS1, yend = MDS2),
  #                arrow = arrow(length = unit(0.03, "npc"))) + 
  geom_text(data = NMDS.spec[NMDS.spec$rank <= maxrank,], 
               aes(x = MDS1, y = MDS2, label = spec))
```

Checking that this represents the default image well (red are species without labels):

```{r}
plot(nmds_run)
```
